{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-09 thesis/related.tex
rephrase & rearrange a few things in related

If I might indulge by listing Landslide in its own related work section \cite{this-thesis},
I would distinguish it by its ability to find shared memory preemption points via dynamic tracing,
rather than relying on user annotations or imprecise compiler instrumentation
as other tools do.
Compared to all other tools I know of,
it implements a wider range of exponential explosion coping techniques,
some theoretical and some heuristic,
some inherited and some novel,
to help the user receive meaningful results as promptly as possible.
Its choice of a familiar pthread-like synchronization API makes it suitable for inexpert users,
and its recent extension to HTM adds support for more modern concurrency patterns as well.

@@

To date a number of techniques have been proposed to mitigate exponential explosion,
the Sisyphean rock of stateless MC.
The notion that some interleavings of concurrent threads could lead to indistinguishable program states and be therefore redundant,
known as {\em partial order reduction} (POR),
was first proposed in \cite{partial-model-checking}
and explored in detail in \cite{partial-order-methods}.
{\em Dynamic POR} (DPOR) was later developed in \cite{dpor},
proposing to track communication events between threads on-the-fly (i.e., dynamically)
rather than to rely on imprecise static alias analyses,
and is now considered the baseline for all subsequent state space reduction approaches in stateless MC.
That paper includes the {\em sleep sets} extension,
which Landslide includes in its implementation.
It is a {\em sound} reduction algorithm, meaning it will never fail to test a possible program behavior, despite skipping many execution sequences.
\cref{sec:landslide-dpor} provides a detailed walk-through of how DPOR works,
as many of this thesis's contributions build directly upon it.
%
DPOR has since been extended in several ways to achieve further reduction
and to incorporate new concurrency models.
\revision{Distributed DPOR \cite{parallel-dpor}
allows the exploration to be parallelized,
with a minimum of overhead from redundant interleavings that would ordinarily be pruned in sequential DPOR.}
Optimal DPOR \cite{optimal-dpor} extends sleep sets into the more expressive {\em wakeup trees},
which provably tests exactly one interleaving from each equivalence class,
i.e., the optimal possible reduction,
at least under the memory independence definition of equivalence.
Extending the equivalence relation itself to capture not just memory {\em address} conflicts
but also the {\em values} read and written,
SATCheck \cite{satcheck} and Maximal Causality Reduction (MCR) \cite{mcr}
use an SMT solver \cite{z3} to identify additional pruning opportunities.
Implementing parallelization, wakeup trees, or SMT-driven exploration in Landslide is left to future work.

@@

Pintos \cite{pintos} has recently emerged as the most popular educational kernel
(by count of top CS schools in the United States who teach by it);
it trades off the prevalence of its concurrency challenges to cover various OS topics more broadly,
especially advanced scheduling algorithms and filesystems.
Pintos is the stand-alone evolution of its predecessor, Nachos \cite{nachos},
which originally ran as a UNIX process with simulated device drivers.
Its popularity motivated me to extend Landslide to support it as an additional kernel architecture
(an unfortunately arduous task)
to prove Landslide's mettle beyond CMU's walls.
Xv6 \cite{xv6}, from MIT, is another major educational kernel, which is also UNIX-like and runs in QEMU,
and a natural target for model checking in future work.
Recently, Columbia introduced a new Android-focused OS course \cite{teaching-android},
which perhaps highlights the importance of related work on model-checking event-driven applications \cite{r4}.

To my knowledge, this is the first study of model checking in an educational setting,
although teaching concurrency is not itself an unstudied problem.
%% tf??
% Eytani et al.~\cite{towards-a-framework} present a promising framework for testing concurrent programs,
% which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
% However, it lacks an evaluation, and makes mention of its educational value only in its future work remarks.
%
%One recent study~
\cite{how-studence} surveyed how students think about
testing and debugging during a concurrent programming project,
finding that unguided, students often approach testing haphazardly,
not understanding the goal of good concurrency coverage,
and also had difficulty understanding single failing executions.
In fact, the study explicitly recommended tool support for testing many thread interleavings automatically (model checking)
and for execution traces to communicate sequences of important events (preemption traces),
which I dare say I have achieved in this thesis.
A more recent study \cite{novices-programmers}
examined in detail the students' thought process
during the diagnosis and fixing phases,
although its participants were drawn from novice-level programming classes,
and the experiment was set up with more elementary bugs like syntax and logic errors correspondingly.
Nevertheless, the authors recommended teaching debugging skills explicitly
via systematic exposure to different kinds of bugs,
which suggests future work for even advanced operating systems curricula
to offer a ``warm up'' Landslide assignment
(for example, the {\tt atomic\_*} tests from \cref{sec:education-pebbles-tests})
that could ultimately lead to a higher solve rate on Landslide's bug reports during P2 (\cref{sec:education-eval-bugs-cmu}).

@@

\revision{The Deadlock Empire~\cite{deadlockempire} is an educational introduction to concurrent programming and debugging,
presented as an online puzzle-hunt of sorts, backed up by a fantasy-themed narrative.
Each puzzle requires the user to manually step through the execution of multiple threads
to find buggy interleavings.
%which in some cases requires the use of an ``expand'' button that
Though itself lacking in scientific analysis of its educational power,
it demonstrates the kind of user-friendly features and overall approach
that many students expressed desire for in their survey responses (\cref{sec:education-eval-survey}).
Future work could extend Landslide's preemption traces to include similar interactivity.}

>>>>

If I might indulge by listing Landslide in its own related work section \cite{this-thesis},
I would distinguish it by its ability to find shared memory preemption points via dynamic tracing,
rather than relying on user annotations or imprecise compiler instrumentation.
%as other tools do.
Compared to all other tools I know of,
it implements a wider range of exponential explosion coping techniques,
some theoretical and some heuristic,
some inherited and some novel,
to help the user receive meaningful results as promptly as possible.
Its choice of a familiar pthread-like synchronization API makes it suitable for inexpert users,
and its recent extension to HTM adds support for more modern concurrency patterns as well.

@@

To date a number of techniques have been proposed to mitigate exponential explosion,
the Sisyphean rock of stateless MC.
The notion that some interleavings
%of concurrent threads
could lead to indistinguishable program states and be therefore redundant,
known as {\em partial order reduction} (POR),
was first proposed in \cite{partial-model-checking}
and explored in detail in \cite{partial-order-methods}.
{\em Dynamic POR} (DPOR) was later developed in \cite{dpor},
proposing to track communication events between threads on-the-fly (i.e., dynamically)
rather than to rely on imprecise static alias analyses,
and is now considered the baseline for all subsequent state space reduction approaches in stateless MC.
That paper includes the {\em sleep sets} extension,
which Landslide includes in its implementation.
It is a {\em sound} reduction algorithm, meaning it will never fail to test a possible program behavior, despite skipping many execution sequences.
\cref{sec:landslide-dpor} provides a detailed walk-through of how DPOR works,
as many of this thesis's contributions build directly upon it.
%
DPOR has since been extended in several ways to achieve further reduction
and to incorporate new concurrency models.
\revision{Distributed DPOR \cite{parallel-dpor}
allows the exploration to be parallelized,
with a minimum of overhead from redundant interleavings that would ordinarily be pruned in sequential DPOR.}
Optimal DPOR \cite{optimal-dpor} extends sleep sets into the more expressive {\em wakeup trees},
which provably tests exactly one interleaving from each equivalence class,
i.e., the optimal possible reduction,
at least under the memory independence definition of equivalence.
Extending the equivalence relation itself to capture not just memory {\em address} conflicts
but also the {\em values} read and written,
SATCheck \cite{satcheck} and Maximal Causality Reduction (MCR) \cite{mcr}
use an SMT solver \cite{z3} to identify additional pruning opportunities.
Implementing parallelization, wakeup trees, or SMT-driven exploration in Landslide is left to future work.

@@

Pintos \cite{pintos} has recently emerged as the most popular educational kernel
(by count of top CS schools in the United States who use it);
it trades off the prevalence of its concurrency challenges to cover various OS topics more broadly,
especially advanced scheduling algorithms and filesystems.
Pintos is the stand-alone evolution of its predecessor, Nachos \cite{nachos},
which originally ran as a UNIX process with simulated device drivers.
Its popularity motivated me to extend Landslide to support it as an additional kernel architecture
(an unfortunately arduous task)
to prove Landslide's mettle beyond CMU's walls.
Xv6 \cite{xv6}, from MIT, is another major educational kernel, which is also UNIX-like and runs in QEMU,
and a natural target for model checking in future work.
Recently, Columbia introduced a new Android-focused OS course \cite{teaching-android},
which perhaps highlights the importance of related work on model-checking event-driven applications \cite{r4}.

To my knowledge, this is the first study of model checking in an educational setting,
although teaching concurrency is not itself an unstudied problem.
%% tf??
% Eytani et al.~\cite{towards-a-framework} present a promising framework for testing concurrent programs,
% which can incorporate model checking as well as static analysis, resource exhaustion, data-race analysis, and coverage analysis.
% However, it lacks an evaluation, and makes mention of its educational value only in its future work remarks.
%
%One recent study~
\cite{how-studence} surveyed how students think about
testing and debugging during a concurrent programming project,
finding that unguided, students often approach testing haphazardly,
not understanding the goal of good concurrency coverage,
and also had difficulty understanding single failing executions.
In fact, the study explicitly recommended tool support
for testing many interleavings automatically (model checking)
and for execution traces to communicate sequences of important events (preemption traces),
which I dare say I have achieved in this thesis.
A more recent study \cite{novices-programmers}
examined in detail the students' thought process
during the diagnosis and fixing phases,
although its participants were drawn from novice-level programming classes,
and the experiment was set up with more elementary bugs like syntax and logic errors correspondingly.
Nevertheless, the authors recommended teaching debugging skills explicitly
via systematic exposure to different kinds of bugs,
which suggests future work for even advanced operating systems curricula
to offer a ``warm up'' Landslide assignment
(for example, the {\tt atomic\_*} tests from \cref{sec:education-pebbles-tests})
that could ultimately lead to a higher solve rate on Landslide's bug reports during P2~(\cref{sec:education-eval-bugs-cmu}).
%
\revision{The Deadlock Empire~\cite{deadlockempire} is an educational introduction to concurrent programming and debugging,
presented as an online puzzle-hunt of sorts, backed up by a fantasy-themed narrative.
Each puzzle requires the user to manually step through the execution of multiple threads
to find buggy interleavings.
%which in some cases requires the use of an ``expand'' button that
Though itself lacking in scientific analysis of its educational power,
it demonstrates the kind of user-friendly features and overall approach
that many students expressed desire for in their survey responses (\cref{sec:education-eval-survey}).
Future work could extend Landslide's preemption traces to include similar interactivity.}

@@
